---
title: "Prediction error with lm()"
author: "Francesco Bailo"
date: "2024-09-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1: Fitting the Model

First, you'll fit the linear model as you did earlier using `lm()`:

```{r}
fit <- lm(formula = mpg ~ wt, data = mtcars)
```

The `formula` of your model, is `mpg ~ wt` which means that for each car model (your observations/rows in your data frame `mtcars`) you are predicting `mpg` (mile per gallons) from `wt` (the weight of the car). 

Your model and your prediction are now stored inside the object `fit`, which contains multiple statistics, including the *coefficients* ($\alpha$ and $\beta$) and the *predicted values* (also known as *fitted values*) for all the cars in the data frame.

## Step 2: Observed values

- **Q**: What is your *observed* value of `mpg` for the car "Ferrari Dino"? There are many ways to access this information stored in the column `mpg` of your data frame.

You can print out the entire data frame 

```{r}
mtcars 
```

But if the data frame is large, it is probably better access the value you need directly, by selecting the column `mpg` and filtering for the row named "Ferrari Dino" (note, that `mtcars` as named columns but also named rows).

```{r}
# Note: use this syntax your_data_frame[ "row name" , "col name" ]
mtcars["Ferrari Dino","mpg"]
```

- **A**: `r mtcars["Ferrari Dino","mpg"]` is the *observed* value for `mpg` of the car "Ferrari Dino"

## Step 3: Predicted values

- **Q**: What is the value of `mpg` predicted from `wt` for "Ferrari Dino" using as predictor?

To extract the *predicted* or *fitted values* from any object created with the function `lm()` you can use `predict()`. Since the data frame `mtcars` had row names, this has also been stored in your object `fit` and you can extract them all, along with the predicted values with:

```{r}
predict(fit)
```
Alternatively, if you want to only extract the value of interest you can

```{r}
# Note: use the [] syntax to extract information from a vector, if the vector 
# is named as in this case you can use a character string.
predict(fit)["Ferrari Dino"]
```

- **A**: `r predict(fit)["Ferrari Dino"]` 

## Step 3: Prediction error

Now we have all the information to compute the prediction error for this car.

Remember that 

$$Error = \text{Observed Value} - \text{Predicted Value}$$ or 

$$\hat{\epsilon} = Y - \hat{Y}$$

```{r}
mtcars["Ferrari Dino","mpg"] - predict(fit)["Ferrari Dino"]
```




