---
title: "Summarise Regression Models with broom"
author: "Francesco Bailo"
date: "2024-09-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The broom package offers you utilities to render regression models objects into the tidy `tibble()`s (a special type of data frame).

## Step 1: Install the broom package (only once)

```{r eval = F}
install.packages("tidymodels")
```

Alternatively, to install just broom

```{r  eval = F}
install.packages("broom")
```

Alternatively, to install just broom with pak

```{r eval = F}
# install.packages("pak") # if you don't have pak already installed
pak::pak("tidymodels/broom")
```

## Step 2: Load the broom package (and the knitr package for `kable()`)

```{r}
library(broom)
```

We also need to load the knitr package to use `kable()` and produce nice looking tables in your PDF or HTML output.

```{r}
library(knitr)
```


## Step 3: Produce a standard regression model table with `broom:tidy()`

```{r}
fit <- lm(Volume ~ Girth + Height, trees)
```

```{r}
tidy(fit) |>
  kable()
```

This table produces all the standard statistics that are generally found in regression tables reporting on a model.

## Step 4: Produce an augemented regression model table with `broom:glance()`

```{r}
glance(fit) |>
    kable()
```

This table produces all the standard statistics that are generally found in regression tables reporting on a model.

```{r}
augment(fit, data = trees) |>
      kable()
```

