---
title: "Week 01</br>Introduction and overview"
subtitle: "SSPS4102 Data Analytics in the Social Sciences<br>SSPS6006 Data Analytics for Social Research<br><br><br>Semester 1, 2026<br>Last updated: `r Sys.Date()`"
author: "Francesco Bailo"
format:
  revealjs:
     # can set self-contained to false if not uploading to Canvas
     # true makes the HTML files pretty big
    self-contained: true
    fig-format: retina
    toc: true
    toc-depth: 1
    toc-title: "In this lecture"
    theme: [default, "assets/sydney.scss"]
    code-line-numbers: false
    slide-number: c
    scrollable: false
    pdf-max-pages-per-slide: 1
    history: false # don't add each slide to browser history
bibliography: assets/pres_bib.bib
csl: assets/apa-old-doi-prefix.csl
execute:
  echo: true
---

## Acknowledgement of Country

I would like to acknowledge the Traditional Owners of Australia and  recognise their continuing connection to land, water and culture. The  University of Sydney is located on the land of the Gadigal people  of the Eora Nation. I pay my respects to their Elders, past and present.

---

```{r, echo = FALSE}
library(tidyverse)
```

# First section

## Quarto

::: callout-note
### Note title

Note text
:::

::: callout-tip
### Tip title

Tip text
:::

::: callout-important
### Important title

Important text
:::

::: notes
Speaker notes go here. Press `s` to see.
:::

::: footer
Footers for comments, or links to further resources. These don't show up when printing the slides to pdf!
:::

## Asides

> Block quote

-   For further details on customisations available see [here](https://quarto.org/docs/presentations/revealjs/)
-   Icons can be included by installing the [fontawesome](https://github.com/quarto-ext/fontawesome) Quarto [extension](https://quarto.org/docs/extensions/).

::: aside
Asides will be on the slide when printed.
:::

# Layouts

## Layout

::: {layout-ncol="2"}
### List One

-   [Item A]{.blue}
-   [Item B]{.red}
-   [Item C]{.purple}
-   Item D

### List Two

-   **Item X**
-   *Item Y*
-   ~~Item Z~~
:::

## Columns (more customisable)

::: columns
::: {.column width="40%"}
Left column
:::

::: {.column width="60%"}
Right column
:::
:::

## Plots: below code

```{r}
library(ggplot2)
p1 = iris |> ggplot() +
  aes(x = Petal.Length, y = Petal.Width, colour = Species) + 
  geom_point(size = 4) + theme_classic(base_size = 20)
p1
```

## Plots: hide code using `echo: false`

```{r}
#| echo: false
p1
```

## Plots: auto two column using `output-location: column`

```{r}
#| output-location: column
#| fig-height: 8
p1 = iris |> ggplot() +
  aes(x = Petal.Length, 
      y = Petal.Width, 
      colour = Species) + 
  geom_point(size = 4) + 
  theme_classic(base_size = 30)
p1
```

## Bullets

When you click the **Render** button a document will be generated that includes:

-   Content authored with markdown
-   Output from executable code

. . .

Reference @tidyverse

## Code

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

## Handy R package

The [`quartostamp`](https://matt-dray.github.io/quartostamp/index.html) R package provides RStudio "Addins" that make life easier for working with Quarto documents. It has a bunch of templates, e.g. for creating

-   callout blocks
-   columns
-   tabsets
-   footnotes
-   pauses
-   speaker notes...

## References
