---
title: "SSPS [4102|6006] Data Analytics</br>[in the Social Sciences|for Social Research]"
subtitle: "Week 08</br>Uncertainty"
author: "Francesco Bailo"
date: "Semester 2, 2024 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "assets/sydney-fonts.css", "assets/sydney.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: true # show a title slide with YAML information
    includes:
      in_header: "assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '4:3' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling

---

background-image: url('assets/USydLogo.svg')
background-size: 95%

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r setup, include=FALSE}

options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      dev = 'svg', 
                      fig.width = 4, 
                      fig.height = 4, out.width="30%",
                      fig.align="center")

library(knitr)
library(kableExtra)
library(tidyverse)
library(sf)
library(DiagrammeR)
library(cowplot)
library(gapminder)

ggplot2::theme_set(theme_bw())

```

---

## Acknowledgement of Country

I would like to acknowledge the Traditional Owners of Australia and recognise their continuing connection to land, water and culture. The  University of Sydney is located on the land of the Gadigal people  of the Eora Nation. I pay my respects to their Elders, past and present.

---

## Recap from last week

- Probability
- Events, Random Variables, and Probability
Distributions
- Probability Distributions
- Bernoulli Distribution
- Normal Distribution
- The Standard Normal Distribution
- Population Parameters vs. Sample Statistics
- Law of Large Numbers and Central Limit Theorem

---

## Today's class

| Time          | Content                              |
|---------------|--------------------------------------|
| 11:00 - 11:10 | Week 08 Student Feedback Survey      |
| 11:10 - 11:30 | Stats: Control/Confounding Variables |
| 11:30 - 11:50 | Stats: Uncertainty (Part 1)          |
| 11:50 - 12:00 | Individual in-class quiz/tutorial    |
| 12:00 - 12:15 | Stats: Uncertainty (Part 2)          |
| 12:15 - 12:30 | Break                                |
| 12:30 - 13:00 | Individual in-class problem set      |
| 13:00 - 14:00 | Group in-class problem set           |

---
class: inverse, center, middle

# Week 08 Anonymous Student Feedback Survey

---
class: inverse, center, middle

# Selecting Control/Confounding Variables in Linear Regression

---

## Why and when we use regression analysis (as social scientists)

- We have variables that capture dimensions that capture interesting concepts (e.g., Democracy, Poverty, Race, Gender, Votes, Education)

- We want to measure the .content-box-yellow[strength and significance of the association] between (generally) two variables: a variable $Y$, which .content-box-yellow[depends] on the value of another variable, $X$. 

This is the relationships we are interested in:

$$X \rightarrow Y$$

And yet the world is a complicated place and many things happen at the same time.

In fact, there are often .content-box-purple[confounding] variables that have an effect on both $X$ and $Y$.

$$Z \rightarrow Y \quad Z \rightarrow X$$

---

## What are Control Variables?

- **Control variables** are additional variables included in your regression model to account for potential confounding factors.
- **Confounders** are variables that are related to both your independent variable(s) and your dependent variable, which can bias the results if not controlled for.
- Including control variables helps .content-box-yellow[isolate] the effect of your main independent variable by adjusting (i.e. **controlling**) for other influences.

---

## Why Include Control Variables?

1. **Reduce Bias**: Get a more accurate estimate of the relationship between your main independent variable and the dependent variable.
2. **Improve Validity**: Make findings more credible and closer to showing a "real" effect.
3. **Clarify Relationships**: Disentangle complex relationships, making it clearer how your main independent variable(s) relate to the dependent variable.

---

## How to Select Control Variables

1. **Theory and Prior Research**: Use existing literature to identify variables known to affect your dependent variable.
2. **Common Sense and Domain Knowledge**: Use your understanding of the subject to identify logical control variables.
3. **Statistical Tests**: Use exploratory data analysis (e.g., correlations) to identify potential controls.
4. **Consistency Across Studies**: Use commonly accepted control variables for comparability.
5. **Avoid Over-Control**: Do not control for mediators, which lie on the causal path between your predictor and the outcome.

---

## Practical Example: Are Private Schools Better Than Public Schools?

- **IV: Private school**: Attending a private school;
- **DP: Test score**: Score in a standard national test.
- **Potential Control/Confounder**: Family wealth, which is likely to affect *both* the likelihood of attending a private school and the test score (e.g., private tutors).
- **What we want to do**: Understand the specific (i.e. net) effect of attending a private school on test scores.
- **Mediator variables**: (Which we should not control for!) "Smaller Class sizes" or "Access to Better Resources"


---

```{r echo = FALSE, out.width = "100%"}
# Load the DiagrammeR package
library(DiagrammeR)

# Create the graph using mermaid syntax
grViz("
  digraph confounding {
    # Define graph attributes
    graph [layout = dot, rankdir = LR]

    # Define nodes
    X [label = 'X\n(I.V.)\nPrivate School\nAtt.', shape = circle, style = 'filled', fillcolor = 'orange']
    Y [label = 'Y\n(D.V.)\nTest Score', shape = circle, style = 'filled', fillcolor = 'orange']
    Z [label = 'Z\nFamily Wealth', shape = circle]
    Z1 [label = 'Z1\nClass size', shape = circle]

    # Define edges (arrows)
    Z -> X [color = black]
    Z -> Y [color = black]
    X -> Y [color = black]
    X -> Z1 [color = black]
    Z1 -> Y [color = black]
  }
")
```

By measuring both $X \rightarrow Y$ and $Z1 \rightarrow Y$, which lies on the path between $X$ or $Y$ you are either attenuating the measurement of what affects test score or hiding it. 

- What affects test score is not anything special with private school, but simply that they offer smaller classes, which is what actually affects students' performance

**Policy lesson**: Don't fund private school, instead reduce the size of classes in public schools. 


---

# On Padlet

.small[

"Does regular exercise lead to better mental health outcomes?"

You measure mental health using a validated scale that assesses anxiety, depression, and overall well-being. You also collect data on how many hours per week each student exercises.

**Variables**:

- **Independent Variable** (IV): Hours of exercise per week
- **Dependent Variable** (DV): Mental health score
= Task: Your task is to identify potential confounders/control variables that could influence both exercise habits and mental health outcomes, potentially biasing the results of the study.

]

.center[

<img src = '../img/padlet_week_08.png' width = '25%'> OR [Link](https://sydney.padlet.org/francescobailo/does-regular-exercise-lead-to-better-mental-health-outcomes-udyw448dn7uq4h54)
]





---

## Key Takeaway

- Selecting control variables is a balance of theory, evidence, and practicality.
- Proper control variables ensure that results reflect the true relationship between predictors and the outcome, without confounding influences.


---
class: inverse, center, middle

# Prof. Llaudet slides on Uncertainty (Part 1)

---
class: inverse, center, middle

# Individual in-class quiz/tutorial

```{r echo = FALSE}
library(countdown)

countdown(minutes = 10, seconds = 00)
```

---
class: inverse, center, middle

# Prof. Llaudet slides on Uncertainty (Part 2)

---
class: inverse, center, middle

# Individual in-class Problem set (with peer-review)

```{r echo = FALSE}
library(countdown)

countdown(minutes = 10, seconds = 00)
```

---
class: inverse, center, middle

# Group in-class problem set

---
class: inverse, center, middle

# Attendance

---
class: inverse, center, middle

# See you next week with Machine Learning!

<iframe src="https://giphy.com/embed/IZY2SE2JmPgFG" width="480" height="274" style="" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/terminator-IZY2SE2JmPgFG">via GIPHY</a></p>
