---
title: "Week 10</br>Count Models and Multilevel Modelling"
subtitle: "SSPS4102 Data Analytics in the Social Sciences<br>SSPS6006 Data Analytics for Social Research<br><br><br>Semester 1, 2026<br>Last updated: `r Sys.Date()`"
author: "Francesco Bailo"
format:
  revealjs:
    self-contained: true
    fig-format: retina
    toc: true
    toc-depth: 1
    toc-title: "In this workshop"
    theme: [default, "assets/sydney.scss"]
    code-line-numbers: false
    slide-number: c
    scrollable: false
    pdf-max-pages-per-slide: 1
    history: false
    html-math-method: katex
bibliography: assets/pres_bib.bib
csl: assets/apa-old-doi-prefix.csl
execute:
  echo: true
---

## Acknowledgement of Country

I would like to acknowledge the Traditional Owners of Australia and recognise their continuing connection to land, water and culture. The University of Sydney is located on the land of the Gadigal people of the Eora Nation. I pay my respects to their Elders, past and present.

## Note

These slides are developed based on:
 
- Alexander, R. (2023). *Telling Stories with Data: With Applications in R*. CRC Press.
- Gelman, A., Hill, J., & Vehtari, A. (2021). *Regression and Other Stories*. Cambridge University Press.

Students are encouraged to refer to the relevant chapters for additional detail and examples.


```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(rstanarm)
library(modelsummary)
library(MASS)
library(broom.mixed)
```

# Review: Bayesian vs Frequentist Approaches (Week 07)

## Two Frameworks for Statistical Inference 

**Recall from Week 7:** There are two main approaches to statistical inference.

. . .

::: columns
::: {.column width="50%"}
### Frequentist Approach

**Probability = Long-run frequency**

- Parameters are **fixed** but unknown
- Data varies across hypothetical repeated samples
- Uncertainty from sampling distribution
- Uses p-values and confidence intervals
- Fast computation with `lm()`, `glm()`
:::

::: {.column width="50%"}
### Bayesian Approach

**Probability = Degree of belief**

- Parameters have **distributions**
- Data is fixed (what we observed)
- Prior beliefs + Data → Posterior beliefs
- Uses credible intervals
- More computational with `rstanarm::stan_lm()`, `rstanarm::stan_glm()`
:::
:::

::: notes
We're using the Bayesian approach this week because it handles complex models well and gives us intuitive probability statements about parameters.
:::

## Key Differences in Practice

| Aspect | Frequentist | Bayesian |
|--------|-------------|----------|
| **Parameter** | Fixed unknown value | Random variable with distribution |
| **Uncertainty** | Confidence interval | Credible interval |
| **Interpretation** | "95% of intervals contain true value" | "95% probability parameter is in interval" |
| **Prior knowledge** | Not incorporated | Incorporated via priors |
| **Computation** | Fast (analytical) | Slower (simulation/MCMC) |
| **R Functions** | `lm()`, `glm()` | `stan_lm()` , `stan_glm()`, `stan_glmer()` |

. . .

::: callout-note
### For This Week
We use `stan_lm()`, `stan_glm()` and `stan_glmer()` for Bayesian estimation. The syntax is similar to `lm()`, glm()`, but the output and interpretation differ.
:::

# Understanding Bayesian Model Output

## Interpreting `stan_lm()` Output: No P-values!

When you run a Bayesian regression, you **don't get p-values**. Instead, you get:

. . .

```{r}
#| echo: false
#| message: false
#| results: hide
set.seed(853)
example_data <- tibble(
  x = rnorm(100, 50, 10),
  y = 2 + 0.5 * x + rnorm(100, 0, 5)
)
example_model <- stan_lm(y ~ x, data = example_data,
                         prior = R2(location = 0.5, what = "mean"), refresh = 0)
```

```{r}
print(example_model)
```

---

**What are we looking at?**

- **(Intercept)**: Median = `r round(coef(example_model)[1], 1)` — Expected y when x = 0
- **x**: Median = `r round(coef(example_model)[2], 2)` — For each 1-unit increase in x, y increases by ~`r round(coef(example_model)[2], 2)` units
- **sigma**: Median = `r round(sigma(example_model), 1)` — Residual standard deviation
- **log-fit_ratio**: Diagnostic for model fit (ignore for now)
- **R2**: Median = `r round(bayes_R2(example_model)[1], 2)` — Proportion of variance explained

::: notes
Notice: No asterisks, no p-values! Instead we have MAD_SD (median absolute deviation of the posterior standard deviation) and credible intervals.
:::

---

```{r}
summary(example_model)
```

---

## What the Bayesian Output Tells You

**Understanding the columns:**

. . .

::: columns
::: {.column width="50%"}
### What You See

- **Median**: Middle value of posterior
- **MAD_SD**: Measure of uncertainty
- **10%**: Lower bound (80% CI)
- **90%**: Upper bound (80% CI)
:::

::: {.column width="50%"}
### What It Means (Frequentist equivalent)

- Like the coefficient estimate
- Like the standard error
- Like confidence interval bounds
- (But with different interpretation!)
:::
:::

. . .

**Example interpretation for x coefficient:**

> "The median estimate is `r round(coef(example_model)[2], 2)`, with an 80% credible interval from the 10% to 90% percentiles. This means there's an 80% probability the true coefficient is in this range."

::: notes
The key difference: Bayesian credible intervals give you direct probability statements about where the parameter is, while frequentist confidence intervals tell you about long-run coverage.
:::

## How to Assess Effects Without P-values

::: callout-important
### Interpreting Uncertainty Without P-values

Instead of asking "Is this effect statistically significant?" (p < 0.05), ask:

1. **Does the credible interval exclude zero?**
   - If the 90% CI doesn't include 0, there's strong evidence for an effect
   - You can also check 95% CI: `posterior_interval(model, prob = 0.95)`

2. **Is the effect size meaningful?**
   - Look at the magnitude of the median coefficient
   - Consider practical significance, not just statistical significance

3. **How certain are we?**
   - Narrower intervals = more certainty
   - MAD_SD tells you about spread of the posterior
:::

. . .

```{r}
#| eval: false
# Get 95% credible intervals
posterior_interval(example_model, prob = 0.95)
```

::: notes
The Bayesian approach shifts focus from binary significance testing to quantifying uncertainty and effect sizes. This is often more informative for research decisions.
:::

# Review: Logistic Regression (Week 9)

## Quick Recap: What is Logistic Regression?

**Since we didn't have the Week 9 seminar**, let's review key concepts about logistic regression:

. . .

**Purpose:** Model **binary outcomes** (0/1, yes/no, success/failure)

$$\Pr(y_i = 1) = \text{logit}^{-1}(\beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \ldots)$$

. . .

**Why not linear regression?**

- Linear regression can predict probabilities outside [0,1]
- Binary outcomes follow a Bernoulli distribution, not Normal
- We need a model that keeps predictions between 0 and 1

::: notes
Logistic regression is fundamental for social science research. Think: voting behavior, policy support, program participation, event occurrence.
:::

## Fitting Logistic Regression in R

**Basic syntax with `glm()`:**

```{r}
#| eval: false
#| echo: true
# Frequentist approach
model <- glm(outcome ~ predictor1 + predictor2,
             data = my_data,
             family = binomial(link = "logit"))
```

. . .

**Bayesian approach with `stan_glm()`:**

```{r}
#| eval: false
#| echo: true
# Bayesian approach
model <- stan_glm(outcome ~ predictor1 + predictor2,
                  data = my_data,
                  family = binomial(link = "logit"))
```

::: callout-note
### Key Argument
`family = binomial(link = "logit")` tells R we're modeling a binary outcome with a logit link function.
:::

## Understanding Log-Odds and Odds Ratios

**Log-odds (logit scale):** The scale on which coefficients are estimated

$$\text{logit}(p) = \log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x_1 + \ldots$$

. . .

**Odds:** The ratio of probability of success to probability of failure

$$\text{Odds} = \frac{p}{1-p}$$

- If $p = 0.5$, odds = 1 (50/50 chance)
- If $p = 0.75$, odds = 3 (3 times more likely to succeed than fail)
- If $p = 0.25$, odds = 0.33 (3 times more likely to fail than succeed)

::: notes
Log-odds can take any value from negative infinity to positive infinity. Odds are always positive. Probabilities are bounded between 0 and 1.
:::

## Interpreting Logistic Regression Coefficients

**Three main approaches:**

. . .

::: {layout-ncol="3"}
### 1. Divide-by-4 Rule

Quick approximation:

$\beta / 4$ = maximum probability change

Example: $\beta = 0.20$
→ max ~5% change in probability

### 2. Odds Ratios

Exponentiate coefficient:

$e^\beta$ = multiplicative change in odds

Example: $\beta = 0.20$
→ $e^{0.20} = 1.22$
→ 22% increase in odds

### 3. Specific Values

Calculate probabilities at specific predictor values:

```{r}
#| eval: false
plogis(b0 + b1*x1)
```
:::

. . .

::: callout-important
### Key Challenge
Unlike linear regression, the effect of a predictor **varies** depending on where you are on the probability scale. Same coefficient → different probability changes.
:::

## Example: Interpreting a Logistic Model

```{r}
#| echo: false
#| message: false
#| results: hide
set.seed(42)
vote_data <- tibble(
  income = sample(1:5, 200, replace = TRUE),
  education = sample(0:1, 200, replace = TRUE),
  vote = rbinom(200, 1, plogis(-1.5 + 0.3*income + 0.5*education))
)
vote_model <- glm(vote ~ income + education, data = vote_data, family = binomial)
```

```{r}
summary(vote_model)$coefficients
```

. . .

**Interpretation:**

- **Income coefficient (0.21):**
  - Divide-by-4: Each income level increases probability by at most 5.2%
  - Odds ratio: $e^{0.21} = 1.23$ → 23% increase in odds of voting

- **Education coefficient (0.55):**
  - Divide-by-4: Education increases probability by at most 13.7%
  - Odds ratio: $e^{0.55} = 1.73$ → 73% increase in odds of voting

# Week 10's content: Count Models and Multilevel Modelling

## Learning Objectives

By the end of this lecture, you will be able to:

- Fit Poisson and negative binomial models for count data
- Understand and diagnose overdispersion
- Introduce multilevel/hierarchical models
- Choose appropriate models for different outcome types
- Understand link functions in GLMs

::: notes
Today we extend our regression toolkit to handle count data - things like number of accidents, number of votes, number of occurrences. We'll also introduce the concept of multilevel modelling.
:::

## This Week's Readings

::: {layout-ncol="2"}
### TSwD

- Ch 13: Generalized linear models
  - 13.3 Poisson regression
  - 13.4 Negative binomial regression
  - 13.5 Multilevel modelling

### ROS

- Ch 15: Other generalized linear models
:::

# Count Data

## What is Count Data?

**Count data** are observations that can only take non-negative integer values: 0, 1, 2, 3, ...

. . .

Examples in social science:

- Number of protest events in a country per year
- Number of children in a household
- Number of times a voter contacts their representative
- Number of crimes reported in a neighbourhood
- Number of A grades awarded in a course

::: notes
Count data is everywhere in social science research. Unlike continuous data, counts have a natural lower bound of zero and can only be whole numbers.
:::

## Why Not Use Linear Regression?

Linear regression assumes:

- Continuous outcome variable
- Normally distributed errors
- Constant variance (homoscedasticity)

. . .

**Problems with count data:**

- Counts are discrete, not continuous
- Cannot be negative (but linear regression can predict negative values!)
- Variance often increases with the mean
- Distribution is often skewed, especially for low counts

::: callout-important
### Key Issue

Using linear regression for count data can produce impossible predictions (like -3.5 crimes) and incorrect standard errors.
:::

## The Poisson Distribution

The **Poisson distribution** is designed for count data. It is governed by a single parameter, λ (lambda), which represents both the mean and variance.

$$P_\lambda(k) = \frac{e^{-\lambda} \lambda^k}{k!}, \text{ for } k = 0, 1, 2, ...$$

. . .

**Key property:** In the Poisson distribution, the mean equals the variance.

$$E(Y) = Var(Y) = \lambda$$

::: notes
Lambda is the expected count. The Poisson distribution is named after French mathematician Siméon Denis Poisson.
:::

## Simulating from the Poisson Distribution

```{r}
#| output-location: column
# Simulate 20 draws with lambda = 3
set.seed(853)
rpois(n = 20, lambda = 3)
```

. . .

```{r}
#| echo: false
#| fig-height: 5
tibble(
  lambda = rep(c(1, 3, 5, 10, 15, 25), each = 10000),
  y = map_dbl(lambda, ~rpois(1, .x))
) |>
  ggplot(aes(x = y)) +
  geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
  facet_wrap(~paste("lambda =", lambda), scales = "free") +
  labs(x = "Count", y = "Frequency",
       title = "Poisson distributions with different λ values") +
  theme_minimal(base_size = 14)
```

::: notes
As lambda increases, the distribution becomes more symmetric and bell-shaped. For small lambda, the distribution is right-skewed.
:::

---

### 📋 Head to Canvas and complete A1 (Section 1-2: Introduction to Count Data and The Poisson Distribution)

<br>

```{r}
#| echo: false
library(countdown)
countdown(
  minutes = 8,
  color_background = "#003366",
  color_text = "#FFD100",
  color_running_background = "#003366",
  color_running_text = "#FFD100",
  color_finished_background = "#FF0000",
  color_finished_text = "white",
  font_size = "3em"
)
```


# Poisson Regression

## The Poisson Regression Model

In Poisson regression, we model the expected count as a function of predictors:

$$y_i \sim \text{Poisson}(e^{X_i\beta})$$

. . .

Or equivalently:

$$\log(\lambda_i) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + ...$$

::: callout-note
### The Log Link Function

We use the **logarithm** as the link function because:

1. It ensures predictions are always positive
2. Coefficients have a multiplicative interpretation
:::

::: notes
The exponential transformation of the linear predictor ensures that lambda is always positive, which makes sense for count data.
:::



## Interpreting Poisson Regression Coefficients

Because we use a log link, coefficients are interpreted as **multiplicative effects** on the expected count.

. . .

For a coefficient β:

- A one-unit increase in x multiplies the expected count by $e^\beta$
- If β = 0.10, then $e^{0.10} = 1.105$, meaning a 10.5% increase
- If β = -0.20, then $e^{-0.20} = 0.82$, meaning an 18% decrease

::: callout-tip
### Quick Approximation

For small coefficients (|β| < 0.1), the coefficient approximately equals the percentage change. For example, β = 0.05 ≈ 5% increase.
:::

## Example: Simulated Poisson Data

```{r}
#| output-location: column
#| fig-height: 6
set.seed(853)
n <- 100
x <- runif(n, -2, 2)
a <- 1
b <- 0.5
linpred <- a + b * x
y <- rpois(n, exp(linpred))
sim_data <- tibble(x = x, y = y)

ggplot(sim_data, aes(x = x, y = y)) +
  geom_point(alpha = 0.6, size = 3) +
  stat_function(
    fun = function(x) exp(a + b * x),
    colour = "red", linewidth = 1
  ) +
  labs(x = "Predictor (x)", 
       y = "Count (y)",
       title = "Simulated Poisson Data") +
  theme_minimal(base_size = 16)
```

::: notes
Notice how the variance increases as the expected count increases - this is characteristic of Poisson data.
:::

## Fitting Poisson Regression in R

Using `glm()` with `family = poisson`:

```{r}
fit_poisson <- glm(y ~ x, family = poisson(link = "log"), data = sim_data)
summary(fit_poisson)
```

## Using rstanarm for Bayesian Estimation

```{r}
#| message: false
#| results: hide
fit_poisson_bayes <- stan_glm(
  y ~ x,
  family = poisson(link = "log"),
  data = sim_data,
  seed = 853
)
```

```{r}
print(fit_poisson_bayes)
```

::: notes
The Bayesian approach gives us uncertainty estimates and works well with default priors for most applications.
:::

## Example: Number of A Grades by Department

Let's simulate data about the number of A grades awarded in university courses across three departments:

```{r}
set.seed(853)
class_size <- 26

count_of_A <- tibble(
  department = c(rep("1", 26), rep("2", 26), rep("3", 26)),
  course = c(paste0("DEP_1_", letters),
             paste0("DEP_2_", letters),
             paste0("DEP_3_", letters)),
  number_of_As = c(
    rpois(n = class_size, lambda = 5),
    rpois(n = class_size, lambda = 10),
    rpois(n = class_size, lambda = 20)
  )
)
```

## Visualising the A Grade Data

```{r}
#| echo: false
#| fig-height: 5
count_of_A |>
  ggplot(aes(x = number_of_As, fill = department)) +
  geom_histogram(position = "dodge", binwidth = 2) +
  labs(x = "Number of As awarded", y = "Number of classes",
       fill = "Department") +
  theme_minimal(base_size = 16) +
  scale_fill_brewer(palette = "Set1")
```

::: notes
We can clearly see the different distributions across departments, with Department 3 awarding the most A grades on average.
:::

## Fitting the Model

```{r}
#| message: false
#| results: hide
grades_model <- stan_glm(
  number_of_As ~ department,
  data = count_of_A,
  family = poisson(link = "log"),
  seed = 853
)
```

```{r}
print(grades_model)
```

## Interpreting the Results

**Coefficients on the log scale:**

- **Intercept (1.32):** Log expected count for Department 1 (baseline)
- **department2 (0.88):** Difference in log expected count between Dept 2 and Dept 1
- **department3 (1.71):** Difference in log expected count between Dept 3 and Dept 1

. . .

**Converting to expected counts:**

```{r}
# Expected counts per department
exp(coef(grades_model)[1])  # Dept 1: exp(1.32) = 3.7 A grades
exp(coef(grades_model)[1] + coef(grades_model)[2])  # Dept 2: exp(1.32 + 0.88) = 9.1 A grades
exp(coef(grades_model)[1] + coef(grades_model)[3])  # Dept 3: exp(1.32 + 1.71) = 20.6 A grades
```

---

**Interpreting as multipliers:**

- Department 2 awards $e^{0.88} = 2.4$ **times as many** A grades as Department 1 (3.7 × 2.4 ≈ 9.1)
- Department 3 awards $e^{1.71} = 5.5$ **times as many** A grades as Department 1 (3.7 × 5.5 ≈ 20.6)

---

### 📋 Head to Canvas and complete A1 (Section 3: Poisson Regression)

<br>

```{r}
#| echo: false
library(countdown)
countdown(
  minutes = 5,
  color_background = "#003366",
  color_text = "#FFD100",
  color_running_background = "#003366",
  color_running_text = "#FFD100",
  color_finished_background = "#FF0000",
  color_finished_text = "white",
  font_size = "3em"
)
```

# Overdispersion

## What is Overdispersion?

**Overdispersion** occurs when the observed variance in the data exceeds what the model predicts.

. . .

For Poisson regression:

- The model assumes $Var(Y) = E(Y) = \lambda$
- If actual $Var(Y) > \lambda$, we have overdispersion

::: callout-important
### Why Does This Matter?

Overdispersion leads to:

- Underestimated standard errors
- Inflated test statistics
- Incorrect p-values
- Overconfident conclusions
:::

::: notes
Overdispersion is extremely common with real-world count data. Almost all real datasets show more variation than the Poisson model predicts.
:::

## Checking for Overdispersion

Compare the mean and variance of your count variable:

```{r}
# For a Poisson distribution, mean ≈ variance
count_of_A |>
  summarise(
    mean = mean(number_of_As),
    variance = var(number_of_As),
    ratio = variance / mean
  )
```

. . .

A ratio much greater than 1 suggests overdispersion.

::: notes
In practice, we rarely see perfect equality between mean and variance. A ratio of 1.5 or less might be acceptable, but larger ratios require attention.
:::

## Causes of Overdispersion

1. **Unobserved heterogeneity:** Important variables are missing from the model

2. **Clustering:** Observations within groups are correlated

3. **Excess zeros:** More zeros than the Poisson distribution predicts

4. **Outliers:** A few extreme values

5. **Wrong distributional assumption:** The data-generating process isn't Poisson

::: notes
Understanding why overdispersion occurs can help you choose the right modelling strategy.
:::

# Negative Binomial Regression

## The Negative Binomial Distribution

The **negative binomial distribution** adds an extra parameter to allow for overdispersion:

$$Var(Y) = E(Y) + \frac{E(Y)^2}{\phi}$$

where φ (phi) is the "reciprocal dispersion" parameter.

. . .

- Lower φ = more overdispersion
- As φ → ∞, the negative binomial approaches the Poisson

::: callout-note
### Flexibility

The negative binomial allows the variance to be larger than the mean, which better fits most real-world count data.
:::

## Simulating Negative Binomial Data

```{r}
#| echo: false
#| fig-height: 5
set.seed(853)
n <- 100
x <- runif(n, -2, 2)
linpred <- 1 + 0.5 * x

sim_nb <- tibble(
  x = rep(x, 3),
  phi = rep(c(0.5, 2, 10), each = n),
  y = c(
    rnegbin(n, exp(linpred), 0.5),
    rnegbin(n, exp(linpred), 2),
    rnegbin(n, exp(linpred), 10)
  )
)

ggplot(sim_nb, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~paste("φ =", phi)) +
  stat_function(fun = function(x) exp(1 + 0.5 * x), colour = "red") +
  labs(title = "Negative Binomial Data with Different Dispersion",
       x = "Predictor", y = "Count") +
  theme_minimal(base_size = 14)
```

::: notes
Notice how the scatter around the curve increases as phi decreases. Lower phi means more overdispersion.
:::

## Fitting Negative Binomial Regression in R

Using `rstanarm`:

```{r}
#| message: false
#| results: hide
# First, let's create some overdispersed data
set.seed(853)
overdispersed_data <- tibble(
  x = runif(100, -2, 2),
  y = rnegbin(100, exp(1 + 0.5 * x), theta = 2)
)

fit_nb <- stan_glm(
  y ~ x,
  family = neg_binomial_2(link = "log"),
  data = overdispersed_data,
  seed = 853
)
```

```{r}
print(fit_nb)
```

## Example: Cause of Death in Alberta, Canada

Real-world example from the textbook: examining leading causes of death in Alberta.

```{r}
#| message: false
#| warning: false
library(janitor)

# Data source: https://open.alberta.ca/opendata/leading-causes-of-death
# Download URL (requires human verification):
# https://open.alberta.ca/dataset/03339dc5-fb51-4552-97c7-853688fc428d/resource/3e241965-fee3-400e-9652-07cfbf0c0bda/download/deaths-leading-causes.csv

alberta_cod <-
  read_csv(
    "../data/deaths-leading-causes.csv",
    skip = 2,
    col_types = cols(
      `Calendar Year` = col_integer(),
      Cause = col_character(),
      Ranking = col_integer(),
      `Total Deaths` = col_integer()
    )
  ) |>
  janitor::clean_names() |>
  add_count(cause) |>
  mutate(cause = str_trunc(cause, 30))
```

::: notes
This dataset contains actual deaths by cause over multiple years in Alberta. We'll model total deaths as a function of cause.
:::

## Exploring the Alberta Data (TSwD, Ch. 13)

```{r}
alberta_cod |>
  filter(
    calendar_year == 2021,
    ranking <= 10
  ) |>
  mutate(total_deaths = format(total_deaths, big.mark = ","))
```

---

```{r}
alberta_cod_top_five <-
  alberta_cod |>
  filter(
    calendar_year == 2021,
    n == 22
  ) |>
  slice_max(order_by = desc(ranking), n = 5) |>
  pull(cause)

alberta_cod <-
  alberta_cod |>
  filter(cause %in% alberta_cod_top_five)
```

---

```{r}
alberta_cod |>
  ggplot(aes(x = calendar_year, y = total_deaths, color = cause)) +
  geom_line() +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Year", y = "Annual number of deaths in Alberta") +
  facet_wrap(vars(cause), dir = "v", ncol = 1) +
  theme(legend.position = "none")
```


---

## Summary Statistics by Cause

```{r}
alberta_cod |>
  group_by(cause) |>
  summarise(
    Min = min(total_deaths),
    Mean = mean(total_deaths),
    Max = max(total_deaths),
    SD = sd(total_deaths),
    Var = var(total_deaths)
) |>
  arrange(desc(Mean)) |>
  head(5) |>
  knitr::kable(digits = 1)
```

---

::: callout-important
### Notice the Variance!
For most causes, **variance > mean**, suggesting overdispersion. This is a hint that negative binomial might be better than Poisson.
:::

---

## Checking for Overdispersion

**Calculating the dispersion ratio:**

$$\text{Dispersion ratio} = \frac{\text{Variance}}{\text{Mean}}$$

```{r}
var(alberta_cod$total_deaths) / mean(alberta_cod$total_deaths)
```

. . .

::: callout-warning
### Severe Overdispersion!
The variance is about **`r round(var(alberta_cod$total_deaths) / mean(alberta_cod$total_deaths), 0)` times larger** than the mean.

- **Poisson assumption:** Variance = Mean (ratio = 1)
- **Our data:** Variance >> Mean (ratio >> 1)
- **Conclusion:** Negative binomial is needed!
:::

## Fitting the Poisson Model

```{r}
#| message: false
#| results: hide
cause_of_death_alberta_poisson <-
  stan_glm(
    total_deaths ~ cause,
    data = alberta_cod,
    family = poisson(link = "log"),
    seed = 853
  )
```

```{r}
print(cause_of_death_alberta_poisson, digits = 2)
```

## Fitting the Negative Binomial Model

```{r}
#| message: false
#| results: hide
cause_of_death_alberta_neg_binomial <-
  stan_glm(
    total_deaths ~ cause,
    data = alberta_cod,
    family = neg_binomial_2(link = "log"),
    seed = 853
  )
```

```{r}
print(cause_of_death_alberta_neg_binomial, digits = 2)
```

::: notes
Notice the reciprocal_dispersion parameter in the negative binomial model - this captures the overdispersion.
:::

## Posterior Predictive Checks: Poisson

**Does the Poisson model fit the data well?**

```{r}
#| fig-height: 4.5
pp_check(cause_of_death_alberta_poisson) +
  theme_minimal(base_size = 14) +
  labs(title = "Posterior predictive check: Poisson model") +
  theme(legend.position = "bottom")
```

::: notes
The light blue lines are simulated datasets from the model. The dark line is the actual data. If the model fits well, the actual data should look like a typical simulation.
:::

## Posterior Predictive Checks: Negative Binomial

**Does the negative binomial model fit better?**

```{r}
#| fig-height: 4.5
pp_check(cause_of_death_alberta_neg_binomial) +
  theme_minimal(base_size = 14) +
  labs(title = "Posterior predictive check: Negative binomial model") +
  theme(legend.position = "bottom")
```

::: notes
The negative binomial model provides a much better fit! The simulated data more closely matches the actual data distribution, especially in capturing the spread.
:::

## Comparing Model Fit

::: columns
::: {.column width="50%"}
### Poisson Model

- Simpler (fewer parameters)
- Assumes mean = variance
- **Poor fit** to Alberta data
- Underestimates variability
:::

::: {.column width="50%"}
### Negative Binomial Model

- More flexible (extra dispersion parameter)
- Allows variance > mean
- **Better fit** to Alberta data
- Captures real-world overdispersion
:::
:::

. . .

::: callout-tip
### Lesson Learned
Real-world count data almost always shows overdispersion. Start with negative binomial to be safe!
:::

## When to Use Which Model?

::: {layout-ncol="2"}
### Use Poisson when:

- Mean ≈ Variance
- Simple count data
- No strong clustering
- Quick preliminary analysis

### Use Negative Binomial when:

- Variance > Mean
- Unexplained heterogeneity
- Real-world count data
- You need accurate standard errors
:::

::: callout-tip
### Rule of Thumb

When in doubt, start with negative binomial. It's safer and reduces to Poisson when overdispersion is minimal.
:::

---

### 📋 Head to Canvas and complete A1 (Sections 4-5: Overdispersion and Negative Binomial Regression)

<br>

```{r}
#| echo: false
library(countdown)
countdown(
  minutes = 8,
  color_background = "#003366",
  color_text = "#FFD100",
  color_running_background = "#003366",
  color_running_text = "#FFD100",
  color_finished_background = "#FF0000",
  color_finished_text = "white",
  font_size = "3em"
)
```

# Exposure and Offset Terms

## What is Exposure?

Sometimes counts occur over different **exposure** periods or populations.

. . .

Examples:

- Number of accidents per intersection (but some intersections have more traffic)
- Number of deaths per region (but regions have different populations)
- Number of events per time period (but observation periods differ)

::: notes
Without accounting for exposure, we might conclude that busier intersections are more dangerous per vehicle, when they might just have more vehicles.
:::

## Modelling Rates with Offset

To model rates rather than raw counts, we use an **offset**:

$$y_i \sim \text{Poisson}(u_i \cdot \theta_i)$$

where $u_i$ is the exposure and $\theta_i = e^{X_i\beta}$ is the rate.

. . .

Taking logs:

$$\log(\lambda_i) = \log(u_i) + X_i\beta$$

The log of the exposure, $\log(u_i)$, is called the **offset**.

## Example: Roach Infestation Study

From ROS Chapter 15: A study of pest management in apartments.

- Outcome: Number of roaches trapped ($y_i$)
- Exposure: Number of trap-days ($u_i$)
- Predictors: Pre-treatment roach level, treatment indicator, building type

```{r}
#| eval: true
library(rstanarm)
data(roaches)

roaches$roach100 <- roaches$roach1/100

fit_roaches <- stan_glm(
  y ~ roach100 + treatment + senior,
  family = neg_binomial_2(link = "log"),
  offset = log(exposure2),
  data = roaches
)
```

::: notes
The offset allows us to model the rate of roaches caught per trap-day, rather than the total count, which would be misleading when trap-days vary.
:::

---

```{r}
summary(fit_roaches)
```

---

## How does the NB fit the data? 

```{r}
pp_check(fit_roaches) +
  scale_x_log10()
```

(Note the log transformation of the x-axis to better visualise the data, which is highgly skewed)

# Introduction to Multilevel Modelling

## The Problem with Complete Pooling

So far, we've treated all observations as independent (complete pooling).

. . .

But data often has **hierarchical structure**:

- Students within schools
- Patients within hospitals
- Voters within electorates
- Time points within individuals

::: callout-important
### Why Does This Matter?

Ignoring grouping structure can lead to:

- Underestimated standard errors
- Incorrect inferences
- Missed important patterns
:::

## Three Approaches

### Complete Pooling

Treat all observations as one group.

*Problem:* Ignores group differences.

### No Pooling

Fit separate models for each group.

*Problem:* Ignores information shared across groups.

### Partial Pooling

Allow groups to differ while "borrowing strength" from other groups.

*Multilevel modelling!*


::: notes
Partial pooling gives us the best of both worlds - it acknowledges group differences while using all available information.
:::

## What is Multilevel Modelling?

**Multilevel models** (also called hierarchical models, random effects models) allow:

- Intercepts to vary by group (varying intercepts)
- Slopes to vary by group (varying slopes)
- Groups to "borrow strength" from each other

. . .

Basic model with varying intercepts:

$$y_{ij} = \beta_0 + \alpha_j + \beta_1 x_{ij} + \epsilon_{ij}$$

where $\alpha_j \sim N(0, \sigma^2_\alpha)$ is the random effect for group $j$.

## Example: Political Support by State

```{r}
#| message: false
#| results: hide
set.seed(853)
political_support <- tibble(
  state = sample(1:10, size = 500, replace = TRUE),
  gender = sample(c(0, 1), size = 500, replace = TRUE),
  noise = rnorm(n = 500, mean = 0, sd = 10),
  supports = as.numeric((state * 2 + gender * 5 + noise) > 20)
)
```

```{r}
#| message: false
#| results: hide
voter_model <- stan_glmer(
  supports ~ gender + (1 | state),
  data = political_support,
  family = binomial(link = "logit"),
  seed = 853
)
```

The `(1 | state)` term specifies varying intercepts by state.

This model allows each state to have its own baseline level of support, while still estimating a common effect of gender.

## Understanding the Output

```{r}
print(voter_model)
```

::: notes
The "Error terms" section shows the variance of the random effects. Larger variance means more variation between groups.
:::

## Visualising Random Effects

```{r}
#| echo: false
#| fig-height: 5
ranef_df <- ranef(voter_model)$state |>
  rownames_to_column("state") |>
  rename(intercept = `(Intercept)`)

ggplot(ranef_df, aes(x = reorder(state, intercept), y = intercept)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "red") +
  coord_flip() +
  labs(x = "State", y = "Random Intercept",
       title = "State-level random effects") +
  theme_minimal(base_size = 16)
```

## When to Use Multilevel Models

Consider multilevel modelling when:

- Data has natural grouping structure
- You want to make inferences about groups
- Sample sizes vary across groups
- Groups are a sample from a larger population

::: callout-tip
### Practical Advice

If you have groups with at least 5-10 observations each and want to account for group-level variation, consider multilevel modelling.
:::

# Summary: Generalised Linear Models

## The GLM Framework

All the models we've covered fit within the **Generalised Linear Model** framework:

1. **Outcome distribution** (data distribution)
2. **Linear predictor** ($X\beta$)
3. **Link function** (connects linear predictor to expected value)

## Common Link Functions

| Distribution | Link | Inverse Link | Use Case |
|--------------|------|--------------|----------|
| Normal | Identity | $\mu = X\beta$ | Continuous data |
| Binomial | Logit | $\mu = \text{logit}^{-1}(X\beta)$ | Binary outcomes |
| Poisson | Log | $\mu = e^{X\beta}$ | Count data |
| Neg. Binomial | Log | $\mu = e^{X\beta}$ | Overdispersed counts |

## Choosing the Right Model

```{r}
#| echo: false
tibble(
  `Outcome Type` = c("Continuous", "Binary (0/1)", "Count (no upper limit)", 
                     "Count (overdispersed)", "Count (with exposure)"),
  Model = c("Linear regression", "Logistic regression", "Poisson regression",
            "Negative binomial", "Poisson/NB with offset"),
  `R Function` = c("lm() or stan_glm()", 
                   "glm(family=binomial) or stan_glm()",
                   "glm(family=poisson) or stan_glm()",
                   "MASS::glm.nb() or stan_glm(family=neg_binomial_2)",
                   "Add offset=log(exposure)")
) |>
  knitr::kable()
```

## Key Takeaways

1. **Count data** requires special models (Poisson or negative binomial)

2. **Overdispersion** is common - the negative binomial handles it well

3. **Offset terms** allow modelling rates when exposure varies

4. **Coefficients** are multiplicative (exponentiate to interpret)

5. **Multilevel models** account for hierarchical data structure

::: callout-important
### Remember

Always check model assumptions and use posterior predictive checks to validate your model choice!
:::

## R Functions Summary

```{r}
#| eval: false
# Poisson regression
glm(y ~ x, family = poisson(link = "log"), data = df)
stan_glm(y ~ x, family = poisson(link = "log"), data = df)

# Negative binomial regression
MASS::glm.nb(y ~ x, data = df)
stan_glm(y ~ x, family = neg_binomial_2(link = "log"), data = df)

# With offset for rates
stan_glm(y ~ x, family = poisson, offset = log(exposure), data = df)

# Multilevel model with varying intercepts
stan_glmer(y ~ x + (1 | group), family = poisson, data = df)
```

## Next Week

**Week 11: Surveys and Experimental Design**

- Principles of randomised controlled trials
- Designing and analysing survey data
- Survey weights and poststratification
- Ethical foundations of experimental research

. . .

**Readings:**

- TSwD Ch 8: Hunt data (8.2-8.4)
- ROS Ch 16-17: Design and poststratification

## Attenande

## Problem set

## References

::: {#refs}
:::

## Additional Reference

Gelman, A., & Vehtari, A. (2024). *Active Statistics: Stories, Games, Problems, and Hands-on Demonstrations for Applied Regression and Causal Inference*. Cambridge University Press. https://doi.org/10.1017/9781009436243
